Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-22 paper/sections/proof-reconstruction.tex
[ paper ] fixed minors.

\node[left = 1.1cm of athena, text width=2cm, align=center]
   (agdafile) {6. \Agda \\ Proof.};

\node[left = 1.1cm of agdafile, text width=2cm, align=center]
   (agda) {7. Proof Checking.};

@@

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system are the proof
assistant, \Agda. We choose \Agda, but a proof assistant with
support of type theory and similar features like \Agda
(see a description of \Agda in Section
\ref{sec:implementation}) can be used.
Following the diagram in Fig.
\ref{fig:proof-reconstruction-workflow}, the process can begin with
a problem in \CPL that we encode in \TPTP format to be sent as the
input of the \Metis prover.
The algorithms of \Metis in Step~3 intend to find
a solution; if they success, \Metis replies a derivation in \TSTP
format of the proof, else the problem is counter feasible.

With such a derivation, we process it with our \Haskell translator
tool, \Athena~\cite{Athena}.
\Athena parses the \TSTP format, analyze the
derivation and generates a representation of the natural deduction
proof using a version of a tree data structure (see the properties
of this tree in Subsection \ref{ssec:metis-proofs}). As a result,
\Athena generates an \Agda file with such a proof. We have included
in this file, \emph{imports} that call \Agda libraries that give
support for the logic~\cite{AgdaProp}, and versions of the \Metis'
inference rules~\cite{AgdaMetis}.

Finally, we use the proof checker in Step~7 to type-check the
proof-term. If type-checking succeed, the proof from such a \TSTP
derivation is correct, else if type-checking fails, one of the three
actors in the process could be the responsible.
They are \Metis, \Athena, or \Agda.

We are responsible for errors committed in the translation from
\TSTP to \Agda: bugs in the parsing module, lack of the coverage of
the \Metis' inference rules since we are emulating them, and last
but not least, the pretty print of the proof in \Agda code.

In the rest of this section, we provide a formal description using
type theory to build definitions and theorems for functions
necessary to emulate \Metis' inference rules.
Some conventions about notations. We will write the
type judgments with ``$a : A$'' to denote that $a$ has type
$\abbre{A}$. Types are written using small caps.
For propositions, we will denote by $φ ≡ ψ$, the syntactic
equivalence.

@@

In the process to emulate \Metis' inference rules in type theory,
we encountered that some of \Metis' transformation for propositions in
the inference rules were given by \emph{general recursion} functions.
But algorithms based on general recursions can not be translated
directly in type theory since it is not a guarantee they terminate.
Then, since a termination checker in type theory recognizes
\emph{structural recursion functions}, we modify our algorithms appropriately.

Using the concept of \emph{structurally smaller} from
\citeauthor{Abel2002}~\cite{Abel2002}. We say, a recursive function is
structurally recursive if it calls itself with only \emph{structurally smaller}
arguments.

To convert a general recursive call into a structurally recursive call, we use
the \emph{Bounded Recursion} technique presented in \cite{Bertot2004} but
others found in \cite{Coquand1992,Abel2002,Bove2005} could be used.
This technique converts a general recursive
function into a structural recursive function by adding an additional
argument that handle the times the function can call itself.
As we show later, we use the complexity of the original function for the new argument in the new function.

For that argument, we add the natural number $n : \Nat$ in the input of the
general recursive function, and perform the recursion over the structure of
\Nat type. \abbre{Nat} type is the usual type with \zero and \suc constructors, and including the common arithmetic operations ($+,\,-,\,*$) and
representations of $\zero, \suc, \suc(\zero), \dots$ with $0,1,2,\dots$.

@@

We define

@@

From the definition above, the size of the argument, ((φ₁ ∧ φ₂) ⇒ φ₃),
is not structurally smaller than the input formula, (φ₁ ⇒ (φ₂ ⇒ φ₃));
((φ₁ ∧ φ₂) ⇒ φ₃) is not a subformula of (φ₁ ⇒ (φ₂ ⇒ φ₃)).

Thus, we have a new function
$\unshunt_{1} : \Prop \to \Nat \to \Prop$
defined by structural recursion in its second argument in
Eq.~\ref{eq:unshunt-structured}.

@@

To feed the second argument of this new function, we will use
functions. For instance, we opted to use a complexity measure of
the formula that $\unshunt_{1}$ function transforms.

A \emph{complexity measure} defined in \cite{Agudelo-Agudelo2017}
and presented in \cite{VanDalen1994} as well as \emph{the rank for a
proposition} is a function with type
$\Prop \to \Nat$ that assigns a natural number to a
proposition φ as follows.

@@

We have already solved the terminating problem for the \unshunt
function in Eq.~\ref{eq:unshunt-structured}, nevertheless we do not
what natural number should be feed that argument in
order to maintain the same outputs as the original \unshunt does in
Eq.~\ref{eq:unshunt-definition}.

Therefore, we define an appropriate complexity measure
$\unshunt_{cm}: \Prop → \NN$
for the second argument of the $\unshunt_{1}$ function in
Eq.~\ref{eq:unshunt-rank-definition}.

@@

Then, a curated version of the $\unshunt_{0}$ function is presented
in Eq.~\ref{eq:unshunt-fixed}.

@@

\emph{Splitting a goal.}
The goal of a proof is processed by \Metis to generated new targets,
\Metis calls these new formulas \emph{subgoals}.
The conjunction of these subgoals is a proposition equivalent to the
original goal. Then, for each subgoal, in a \Metis' \TSTP derivation
we find a refutation.

@@

In the following, we show how the $split$ function jointly with Theorem.~\ref{thm:thm-strip} prove the correctness of the process mentioned above.
First, we need to prove the correctness for the $\unshunt$ function in Eq~\ref{eq:unshunt-fixed}.

@@

\begin{proof} Use induction on the cases defined by the outcome of the
$\unshunt$ function in Eq.~\ref{eq:unshunt-structured}.
When $n = 0$, by definition, $Γ ⊢ φ$.
When $n = 1$, we apply the following theorem presented in
\cite{AgdaProp}

@@

Now, let us consider the $split$ function that
yields the conjunction of subgoals from a proposition goal defined in Eq~\ref{eq:split-definition}.

>>>>

\node[left = 1.1cm of athena, text width=2cm, align=center]
   (agdafile) {6. Proof-term.};

\node[left = 1.1cm of agdafile, text width=2cm, align=center]
   (agda) {7. Proof Checker.};

@@

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is a proof
assistant, \Agda. We choose \Agda, but another proof assistant with the same support of type theory and inductive types is valid (see a description of \Agda for its main features in
Section~\ref{sec:implementation}).
The overview of the proof-reconstruction is presented in
Fig.~\ref{fig:proof-reconstruction-workflow}, the process can begin with
a problem in \CPL that we encode in \TPTP format to be sent as the
input of the \Metis prover.
The algorithms of \Metis in Step~3 find
a solution if the problem is a theorem, in that case \Metis replies
with a derivation in \TSTP format of the proof, otherwise the problem is counter feasible.

With such a derivation, in Step~5, we process the derivation with our \Haskell translator tool, \Athena~\cite{Athena}.
\Athena parses the \TSTP format, analyze the
derivation and generates a representation of the natural deduction
proof using a tree data structure (see the properties
of this tree in Subsection \ref{ssec:metis-proofs}).
We perform an analysis of the proof-tree removing redundancies,
unnecessary steps or unused output from the \TSTP derivation.

As a result,
\Athena generates an \Agda file with the proof-term. We have included
in this file, \emph{imports} that call our \Agda libraries that give
support for CPL presented in~\cite{AgdaProp}, and support for \Metis'
inference rules implemented in~\cite{AgdaMetis}.

Finally, we use the proof checker in Step~7 to type-check the
proof-term. If type-checking succeed, the \TSTP
derivation is correct, otherwise when it fails, one of the three
actors in the process could be the responsible: \Metis, \Athena, or
\Agda.

We are responsible for errors committed in the translation from
\TSTP to \Agda: bugs in the parsing module, lack of the coverage of
\Metis' inference rules, and last but not least, the pretty print of
the proof in \Agda code.

In the rest of this section, we provide a formal description using
type theory to build definitions and theorems for functions
necessary to emulate \Metis' inference rules.
Some conventions about notations. We will write the
type judgments with ``$a : A$'' to denote that $a$ has type
$\abbre{A}$. Types are written using small caps.
For propositions, we will denote by $φ ≡ ψ$, the syntactic
equivalence. We use the \Prop type for propositions and for theorems,
the dependent type, $ Γ ⊢ φ$ where $Γ$ has type $\abbre{List} \Prop$ and
$φ$ has type $\Prop$.

@@

To emulate \Metis' inference rules in type theory,
we observed that transformations performed by some rules were given by \emph{general recursive} functions.
But algorithms based on general recursions can not be translated
directly in type theory since it is not a guarantee they terminate.

Then, we solve the termination problem by modifying problematic functions
to be \emph{structural recursive}~\cite{Coquand1992,Abel2002,Bove2005}

We use the same notion of \emph{structurally smaller} from
\citeauthor{Abel2002}~\cite{Abel2002} to say, a recursive function is
structurally recursive if it calls itself with only
\emph{structurally smaller}
arguments.

To convert a general recursive call into a structurally recursive call,
we use the \emph{Bounded Recursion} technique presented
in~\cite{Bertot2004} but
others found in \cite{Coquand1992,Abel2002,Bove2005} could be used.
This technique converts a general recursive
function into a structural recursive function by adding an additional
argument for the function complexity, the bound for the number of times, the function needs to call itself to get the final result.

Then, we add the natural number $n : \Nat$ in the input of the
general recursive function, and perform the recursion over the structure of
\Nat type. \abbre{Nat} type is the usual type for natural numbers with \zero and \suc as their constructors. We consider the common arithmetic operations ($+,\,-,\,*$) and numbers $0,1,2,\dots$ as representations of $\zero, \suc, \suc(\zero), \dots$.

@@

@@

From the definition above, the size of the argument, ((φ₁ ∧ φ₂) ⇒ φ₃),
is not structurally smaller than the input formula, (φ₁ ⇒ (φ₂ ⇒ φ₃)), since ((φ₁ ∧ φ₂) ⇒ φ₃) is not a subformula of (φ₁ ⇒ (φ₂ ⇒ φ₃)).

Thus, we define the function
$\unshunt_{1} : \Prop \to \Nat \to \Prop$
by structural recursion in its second argument in
Eq.~\ref{eq:unshunt-structured}.

@@

As we mentioned above, the natural number in the second argument of the function used in Eq.~\ref{eq:unshunt-structured} stands for a complexity measure of the function, in this case $\unshunt_{1}$ function.

A \emph{complexity measure} defined in \cite{Agudelo-Agudelo2017}
and presented in \cite{VanDalen1994} as \emph{the rank for a
proposition}, is a function $\Prop \to \Nat$ that assigns a natural number for a proposition φ as follows.

@@

Recall, we want to maintain the same outputs as the original \unshunt does in Eq.~\ref{eq:unshunt-definition} and the complexity measure of it, is a bound for the number of times this function can call itself.

Therefore, we define an appropriate complexity measure
$\unshunt_{cm}: \Prop → \NN$ for our $\unshunt_{1}$ function presented in
Eq.~\ref{eq:unshunt-rank-definition}.

@@

Finally, the curated version of $\unshunt_{0}$ function is presented
in Eq.~\ref{eq:unshunt-fixed}.

@@

\emph{Splitting a goal.}
The goal of a problem to be proved by \Metis is processed to generate new
propositions, the new targets, \Metis calls these formulas
\emph{subgoals}.

The conjunction of these subgoals is a proposition equivalent to the
original goal. Then, we find a refutation for each subgoal in the \Metis' \TSTP derivation.

@@

In the following, we show how Theorem.~\ref{thm:thm-strip} proves the correctness of the process mentioned above. But first, let prove some lemmas for $\unshunt$ function presented in Eq.~\ref{eq:unshunt-fixed},
and $split$ function from the definition in Eq.~\ref{eq:split-definition}.

@@

\begin{proof} Use induction on the cases defined by the outcome of the $\unshunt$ function in Eq.~\ref{eq:unshunt-structured}.
When $n = 0$, by definition, $Γ ⊢ φ$.
When $n = 1$, we apply the following theorem from~\cite{AgdaProp}

@@

Now, let us consider the $split$ function presented in
Eq.~\ref{eq:split-definition} that yields a conjunction of subgoals from a proposition, the goal in \Metis.
